Making Sequential Programs Deterministic

### 2019-03-28 design.tex
Edits in section 2.4

In a type system,
a \textit{type} stands for a set of values.  It abstracts or restricts the
set of possbile run-time values that an expression may evaluate to.
A programming language provides \emph{basetypes}, such as \<int>.
A \textit{type qualifier} on a \textit{type} adds additional constraints;
that is, it reduces the size of the set of values.
A type checker can verify the correctness of this \textit{type} with its dataflow analysis and subtyping rules.
In Java, a type qualifier is written as a type annotation with a leading \<@>, as in
\<@Positive int>.

@@

The core of the determinism type system is the following type qualifiers (see Figure~\ref{fig-determinism-hierarchy}):
%\todo{introduce just @Det and @NonDet first, and then @OrderNonDet after polymorphism.}
\begin{itemize}
    \item \<NonDet> indicates
    that the expression might have different values in two different executions.
    \item \<Det> indicates that
    the expression evaluates to the same value (with respect to \<.equals()>) in all
    executions; for a collection, iteration also yields the values in the same
    order.
\end{itemize}

\subsection{Handling order nondeterminism}
In addition to the type qualifiers described in~\ref{type-hierarchy}, \theDeterminismChecker 
defines the type qualifier \<OrderNonDet>. This qualifier represents
a collection or an array that will have the same elements in every execution, but in a
possibly different order.  \<OrderNonDet> may only be written on
collections and arrays.
A collection is any subtype of \<java.util.Collection>,
\<java.util.Iterator>, or \<java.util.Map>, including user-defined types. The relationship
of \<OrderNonDet> with respect the the other two type qualifiers is \codeid{Det $\sqsubseteq$ OrderNonDet $\sqsubseteq$ NonDet} as shown in Figure~\ref{fig-determinism-hierarchy}.

@@

@@

    $\infer[\textsc{ordernondet collection}]{\wellformed{\kappa_c \  \beta_c \langle\kappa_e \ \beta_e\rangle}}{\kappa_c : \|OrderNonDet/Det| & \kappa_e \sqsubseteq \kappa_c & \beta_c \sqsubseteq \|Collection/Iterator|}$

@@

    $\infer[\textsc{nondet collection}]{\wellformed{\|NonDet| \ \beta_c\langle\|NonDet| \ \beta_e\rangle}}{\beta_c \sqsubseteq \|Collection/Iterator|}$

@@

     $\infer[\textsc{ordernondet array}]{\wellformed{\kappa_e \ \beta_e \ \  \kappa_a []}}{\kappa_a : \|OrderNonDet/Det| & \kappa_e \sqsubseteq \kappa_a }$

@@

    $\infer[\textsc{nondet array}]{\wellformed{\|NonDet| \ \beta_e \ \  \|NonDet| []}}{}$

    $\infer[\textsc{ordernondet map}]{\wellformed{\kappa_m \  \beta_m \langle\kappa_k \ \beta_k, \kappa_v \ \beta_v\rangle}}{\kappa_m: \|OrderNonDet/Det| & \kappa_k \sqsubseteq \kappa_m & \kappa_v \sqsubseteq \kappa_m & \beta_m \sqsubseteq \|Map|}$

    $\infer[\textsc{nondet map}]{\wellformed{\|NonDet| \ \beta_m\langle\|NonDet| \ \beta_k, \|NonDet| \ \beta_v\rangle}}{\beta_m \sqsubseteq \|Map|}$

    \caption{Type well formedness}
    \label{type-validity}
\end{figure}

Figure~\ref{type-validity} shows the type validity rules of our type system. A type $\tau$ is represented as $\kappa \ \beta$
where $\kappa$ is a type qualifier from the determinism type hierarchy and $\beta$ is a Java type.
As stated in the rule $\textsc{any type}$, the types \<Det> and \<NonDet> can be written on any Java type.
The (determinism) type of a Collection or Iterator must be a supertype or equal to
the type of the type parameter (Rule $\textsc{ordernondet collection}$) (see Figure~\ref{fig-determinism-collections}).
If the Collection is \<NonDet>, then the type parameter may not be
\<Det> or \<OrderNonDet> (Rule $\textsc{nondet collection}$ in Figure~\ref{type-validity}). Although such types could exist, they are
disallowed to prevent the following bug:

@@

Some examples of valid types are:
\begin{itemize}
    \item \codeid{NonDet\ \ \ \ \ \ List<NonDet\ \ \ \ \ \ Integer>}
    \item \codeid{Det\ \ \ \ \ \ \ \ \ List<Det\ \ \ \ \ \ \ \ \ Integer>}
    \item \codeid{OrderNonDet List <OrderNonDet Set<...>\relax >}
    \item \codeid{OrderNonDet List <Det String>}
\end{itemize}

These types are invalid:
\begin{itemize}
    \item \codeid{OrderNonDet Set <NonDet\ \ \ \ \ \ Integer>}
    \item \codeid{NonDet\ \ \ \ \ \ List<Det\ \ \ \ \ \ \ \ \ Integer>}
    \item \codeid{Det\ \ \ \ \ \ \ \ \ List<OrderNonDet Set<...>\relax >}
\end{itemize}

Similarly, the (determinism) type of an array must be a supertype or equal to
the type of the component type (see Figure~\ref{fig-determinism-collections}).
As with collections, \codeid{NonDet} arrays of \codeid{Det} or \codeid{OrderNonDet}
elements are not allowed (Rules $\textsc{ordernondet array}$ and $\textsc{nondet array}$ in Figure~\ref{type-validity}).

\begin{figure}
    \centering
    \begin{tabular}{|l|l|l|l|l|}
        \cline{3-5}
        \multicolumn{2}{c|}{~}  &  \multicolumn{3}{c|}{Type Argument (or array component type)} \\ \cline{3-5}
        \multicolumn{2}{c|}{~}  & NonDet     & OrderNonDet & Det \\ \hline
        & NonDet      &   valid    &  invalid    & invalid  \\ \cline{2-5}
        Collection (or array)   & OrderNonDet &   invalid  &  valid      & valid  \\ \cline{2-5}
        & Det         &   invalid  &  invalid    & valid      \\ \hline
    \end{tabular}
    \caption{Valid Collection (and array) declarations.  The Collection's (or array's) type qualifier
        must be a supertype or equal to the type argument (or array component type).}
    \label{fig-determinism-collections}
\end{figure}

A collection (\<List, Set, or a Map>) annotated as \<OrderNonDet> has the following properties:
\begin{enumerate}
    \item The individual elements extracted by iterating over the collection get the type \<NonDet>.
    \item Accessing properties sich as \<size()> or querying if the collection \<isEmpty()> will return types
    that are annotated as \<Det> as they do not depend on the iteration order. 
\end{enumerate}

A user defined type may be annotated as \<OrderNonDet> if and only if it
is a subtype of Collection or Iterator.
For example, if a user defines a type as
\begin{verbatim}
public class TestUserCollection<E> extends ArrayList<E> {...}
\end{verbatim}
Writing \codeid{OrderNonDet TestUserCollection<Det Integer>} is valid.\\
Writing \codeid{OrderNonDet TestUserCollection<NonDet Integer>} is invalid
because the type of the type parameter (\codeid{NonDet}) is not a subtype
of the user defined Collection type (\codeid{OrderNonDet}).

User defined types that are not a subtype of Collection or Iterator
may not be annotated as \codeid{OrderNonDet}.

@@

Recall that \theDeterminismChecker does not permit \<OrderNonDet> on a type that isn't a \<Collection>, \<Iterator>, or an array.
As a consequence, resolution of polymorphic type qualifiers at method signatures could have undesirable effects.
Suppose a method declaration is annotated as \codeid{PolyDet int size(PolyDet List<T> list)}. Invoking this method
with an argument of type \codeid{OrderNonDet List<Det String>} will result in \<PolyDet> getting resolved to \<OrderNonDet> which is invalid
on the return type \<int>. We circumvent this scenario by introducing two variants to the \<PolyDet> qualifier, \<PolyDet$\uparrow$>
and \<PolyDet$\downarrow$>.

@@

\begin{figure}
    $\infer[\textsc{poly resolution}]{\pi_r \ \beta_r method(\overline{\pi_i \ \beta_i})\{\ \}}{\forall K \in \{\|NonDet, OrderNonDet, Det|\} : K \ \beta_r method(\overline{K \ \beta_i})\{\ \} & \pi_r, \pi_i \in \{\|PolyDet|, \|PolyDet|\uparrow, \|PolyDet|\downarrow    \}}$
    \caption{Polymorphic resolution rule}
    \label{fig-poly-resolution}
\end{figure}

\begin{table}[]
    \begin{tabular}{|l|l|l|}
        \hline
        \textbf{\<PolyDet>} & \textbf{\<PolyDet$\uparrow$>} & \textbf{\<PolyDet$\downarrow$>} \\ \hline
        \<NonDet> & \<NonDet> &  \<NonDet>\\ \hline
        \<OrderNonDet> & \<NonDet> &  \<Det>\\ \hline
        \<Det> & \<Det> &  \<Det>\\ \hline
    \end{tabular}
\caption{Polymorphic resolution}
\label{tab-poly-resolutions}
\end{table}

@@

In order to provide soundness guarantees, we must ensure that a collection that is
declared as \<Det> is not side-effected in non-deterministic ways. Consider the following qualifier 
on the List add method: \codeid{public void add(PolyDet ArrayList<E> this, PolyDet int index, E element)}.
Suppose a client calls this method as follows:
\begin{verbatim}
@Det ArrayList<@Det String> list;
@NonDet int random;
@Det String str;
list.add(random, str);
\end{verbatim}
This will result in the instantion \codeid{public void add(@NonDet ArrayList<E> this, @NonDet int index, E element)} and
the method invocation will type check. This is problematic because even though \<list> was declared to be \<Det>,
the type checker allows the addition of an element at a \<NonDet> location violating our determinism guarantees.
We prevent this behavior by introducing another variant of \<PolyDet>, \<PolyDet("use")> that does not affect
type instantiations. This is especially useful in preventing methods from non-deterministically modifying the state
of a deterministic receiver. To have this effect, the library method \<ArrayList.add> is annotated as:
\codeid{public void add(PolyDet ArrayList<E> this, PolyDet("use") int index, E element)}.
Since \<PolyDet("use")> has no effect on type instantiations, calling this method on a \<Det List>
will result in the method declaration being resolved as \codeid{public void add(@Det ArrayList<E> this, @Det int index, E element)}
irrespective of the type of \<index>, thereby preventing undesired side-effects.

@@

With the above rules, the checker framework will automatically annotate a local collection with type \<NonDet> and its
type arguments will get the type qualifier \<Det>. For example, a locally declared \codeid{List} of \codeid{Strings} will be automatically annotated
as \codeid{NonDet List<Det String>}. Recall from~\ref{collection-rules} that \theDeterminismChecker does not allow this type on
any collection. While this is still sound, it will result in \theDeterminismChecker reporting a lot of errors because every locally declared 
collection gets annotated with this invalid type. To avoid this situation, we make an exception to the defaulting rules for local collections
and annotate their type parameters also with \<NonDet>. So a local list of Strings gets the default qualifier of \codeid{NonDet List<NonDet String>}. While this defaulting eliminates the possibility of automatically annotating locals collections with an invalid
type, it could result in a high number of false positives. In the following code snippet,
\begin{verbatim}
void test(@Det List<@Det String> argList) {
    List<String> localList = argList;
}    
\end{verbatim}
\theDeterminismChecker would report an error at the assignment statement since the inferred type for \codeid{localList} is 
\codeid{NonDet List<NonDet> String} and since collections types are invariant.
We therefore recommend that programmers explicitly annotate
all local variables that are collections to reduce false positives. The same reasoning and recommendation applies to arrays.

\TheDeterminismChecker has additional type refinement rules for the sorting methods as shown in Figure~\ref{fig-sorting}. 
\begin{figure}
    $\infer[\textsc{list sort}]{\Gamma \vdash x : \|Det \ List|\langle\|Det| \ \beta\rangle}{\Gamma \vdash x : \|OrderNonDet \ List|\langle\|Det| \ \beta\rangle & \|x.sort()|}$

    $\infer[\textsc{collections sort}]{\Gamma \vdash x : \|Det \ List|\langle\|Det| \ \beta\rangle}{\Gamma \vdash x : \|OrderNonDet \ List|\langle\|Det| \ \beta\rangle & \|Collections.sort.sort(x)|}$

    $\infer[\textsc{arrays sort}]{\Gamma \vdash x : \|Det \ \beta \ Det[]| }{\Gamma \vdash x : \|Det \ \beta \ OrderNonDet[]| & \|Arrays.sort( \ x)|}$

    \caption{Sorting type refinement rules}
    \label{fig-sorting}
\end{figure}
\TheDeterminismChecker type refines a receiver of type \codeid{OrderNonDet List} to a \codeid{Det List} if all the
type argument of this \codeid{List} have the type qualifier \codeid{Det} (Rule LIST SORT in Figure~\ref{fig-sorting}). Similar rules apply for \codeid{Arrays.sort()} and \codeid{Collections.sort()}.
As per these rules, if \codeid{sort()} is called on a receiver of type \codeid{OrderNonDet List<Det Set<Det String>>}, it will
be type refined to \codeid{Det List<Det Set<Det String>>}. However a receiver of type \codeid{OrderNonDet List<OrderNonDet Set<Det String>>} will not be type refined.

@@

In the JDK, \<LinkedHashSet> is implemented as a subclass of \<HashMap>.
\todo{Talk about this design choice. Josh Blochs book: "If the code surrounding the first declaration depended on LinkedHashSet's ordering policy,
    then it would be incorrect to substitute HashSet for LinkedHashSet in the declaration, because HashSet makes no guarantee concerning iteration order."}

\subsection{Determinism checker specific rules for precision}\label{precision}
We annotated the \codeid{equals()} method of the \<Set> class in the JDK as
\begin{verbatim}
PolyDet$\uparrow$ boolean equals(PolyDet Set<E> this, PolyDet Object o)
\end{verbatim}
While this is sound, it is imprecise in the case of \codeid{OrderNonDet Sets}. We added the following rule shown in Figure~\ref{fig-precision-rules} to our checker to improve precision.
\begin{figure}
    \todo{linebreak}
    $\infer[\textsc{set precision}]{\Gamma \vdash \|z : Det|}{\Gamma \vdash \|x| : \|OrderNonDet| \ \|Set|\langle \kappa_{x} \ \beta_{x}\rangle, \|y| : \|OrderNonDet| \ \|Set|\langle \kappa_{y} \ \beta_{y}\rangle & \|z = x.equals(y)| & \langle \kappa_{x} \ \beta_{x}\rangle \nsqsubseteq \langle \|OrderNonDet \ List| \rangle & \langle \kappa_{y} \ \beta_{y}\rangle \nsqsubseteq \langle \|OrderNonDet \ List| \rangle }$

@@

    \caption{Improving precision}
    \label{fig-precision-rules}
\end{figure}
According to this rule, if the \<Set> receiver has type \codeid{@OrderNonDet Set<@Det List} and the argument to the \codeid{equals()} method is
also of the same type, the return type gets the qualifier \<Det>. Without this precision rule, the return type would have resolved to \<NonDet> which
is sound but imprecise. This rule is careful enough to not annotate the return type of \<equals()> as \<Det> if a type argument to either the
receiver or the argument happens to be an \codeid{OrderNonDet List} (An order nondeterminstic list does not provide any guarantee on the
equality as it is dependent on the iteration order).

@@

>>>>

In a type system,
a \textit{type} stands for a set of values.  It abstracts or restricts the
set of possible run-time values that an expression may evaluate to.
A programming language provides \emph{basetypes}, such as \<int>.
A \textit{type qualifier} on a \textit{type} adds additional constraints;
that is, it reduces the size of the set of values.
A type checker can verify the correctness of this \textit{type} with its dataflow analysis and subtyping rules.
In Java, a type qualifier is written as a type annotation with a leading \<@>, as in
\<@Positive int>.

@@

The core of the determinism type system is the following type qualifiers (see Figure~\ref{fig-determinism-hierarchy}):
%\todo{introduce just @Det and @NonDet first, and then @OrderNonDet after polymorphism.}
\begin{itemize}
    \item \<NonDet> indicates
    that the expression might have different values in two different executions.
    \item \<OrderNonDet> indicates that the expression is a collection or
      map (dictionary or associative array) that
      will have the same elements in every execution, but in a possibly different order.
    \item \<Det> indicates that
    the expression evaluates to the same value (with respect to \<.equals()>) in all
    executions; for a collection, iteration also yields the values in the same
    order.
\end{itemize}

\<OrderNonDet> may only be written on collections and maps.  Our implementation
handles array types, and any class (including user-defined classes) that
implements the \<Iterable> or \<Iterator> interfaces; this includes all
Java collections such as \<List>s and \<Set>s.  For simplicity, and because
the semantics and rules are similar even though the syntax varies, this
paper uses the term ``collection'' and the type \<Collection> to represent
all these types.

Our type system largely treats a map as a collection of key--value pairs.  Like any
collection, a map may be \<Det>, \<OrderNonDet>, or \<NonDet>.

@@

\todo{Note that the \textsc{ordernondet} versions also handle \<Det>,
  because $\<Det> \sqsubseteq \<OrderNonDet>$.}

@@

    $\infer[\textsc{ordernondet collection}]{\wellformed{\kappa_c \  \beta_c \langle\kappa_e \ \beta_e\rangle}}{\kappa_c : \|OrderNonDet/Det| & \kappa_e \sqsubseteq \kappa_c & \beta_c \sqsubseteq \<Collection>}$

@@

    $\infer[\textsc{nondet collection}]{\wellformed{\|NonDet| \ \beta_c\langle\|NonDet| \ \beta_e\rangle}}{\beta_c \sqsubseteq \<Collection>}$

@@

%      $\infer[\textsc{ordernondet array}]{\wellformed{\kappa_e \ \beta_e \ \  \kappa_a []}}{\kappa_a : \|OrderNonDet/Det| & \kappa_e \sqsubseteq \kappa_a }$
%     
%     \bigskip
%     
%     $\infer[\textsc{nondet array}]{\wellformed{\|NonDet| \ \beta_e \ \  \|NonDet| []}}{}$
%     
%     \bigskip

    $\infer[\textsc{ordernondet map}]{\wellformed{\kappa_m \  \beta_m \langle\kappa_k \ \beta_k, \kappa_v \ \beta_v\rangle}}{\kappa_m: \|OrderNonDet/Det| & \kappa_k \sqsubseteq \kappa_m & \kappa_v \sqsubseteq \kappa_m & \beta_m \sqsubseteq \<Map>}$

@@

    $\infer[\textsc{nondet map}]{\wellformed{\|NonDet| \ \beta_m\langle\|NonDet| \ \beta_k, \|NonDet| \ \beta_v\rangle}}{\beta_m \sqsubseteq \<Map>}$

    \caption{Type well formedness}
\todo{The \textsc{any type} rule is not correct, because not every type is
  well-formed.  It needs an antecedent stating that $\beta$ is not a
  collection or map.}

\todo{The collection rules do not permit nested collections; similarly, the
  (currently commented out) array rules did not permit any two-dimensional
  array to be well-formed.  There is a bigger issue here.
  Consider the ``nondet collection'' rule.  $\beta_c$ is a class (we need
  to use a different symbol for it, probably $c$), but $\beta_e$ should be
  a full type, which might contain type qualifiers within it.  (Actually,
  $\|NonDet|\ \beta_e$ is a type, not $\beta_e$ itself.)  So, there should
  be an antecedent stating that $\|NonDet| \beta_e$ is well-formed, and
  for the ``ordernondet collection'' rule stating that $\kappa_e \beta_e$
  is well-formed.  Likewise, antecedents are needed for the map rules.
  Otherwise, these rules permit construction of $\<Map>\langle\tau'\rangle$
  where $\tau'$ is itself not well-defined.}

\todo{``$\kappa : \|Det / NonDet|$'' should be ``$\kappa = \|Det /
  NonDet|$'' or (better)
  ``$\kappa \in \{\|Det|, \|NonDet|\}$''.
  You could also express $\kappa_c : \|OrderNonDet/Det|$ as just
  $\kappa_c \sqsubseteq \|OrderNonDet|$ since $\|Det| \sqsubseteq \|OrderNonDet|$.
}

  \label{type-validity}
\end{figure}

Figure~\ref{type-validity} shows the type validity\todo{The caption says
  ``type well formedness''.  Choose one term and use it consistently
  throughout.}
rules of our type system. A type $\tau$ is represented as $\kappa \ \beta$
where $\kappa$ is a type qualifier from the determinism type hierarchy and $\beta$ is a Java type.
As stated in the rule $\textsc{any type}$, the types \<Det> and \<NonDet> can be written on any Java type.
The (determinism) type of a collection must be a supertype or equal to
its element type (Rule $\textsc{ordernondet collection}$) (see Figure~\ref{fig-determinism-collections}).
If the Collection is \<NonDet>, then the type parameter may not be
\<Det> or \<OrderNonDet> (Rule $\textsc{nondet collection}$ in Figure~\ref{type-validity}). Although such types could exist, they are
disallowed to prevent the following bug:

@@

Some examples of valid types are:\todo{double-check}
\begin{itemize}
    \item \codeid{NonDet\ \ \ \ \ \ List<NonDet\ \ \ \ \ \ Integer>}
    \item \codeid{OrderNonDet List<OrderNonDet Set<...>\relax >}
    \item \codeid{OrderNonDet List<Det Integer>}
    \item \codeid{Det\ \ \ \ \ \ \ \ \ List<Det\ \ \ \ \ \ \ \ \ Integer>}
\end{itemize}

These types are invalid:
\begin{itemize}
    \item \codeid{NonDet\ \ \ \ \ \ List<OrderNonDet Integer>}
    \item \codeid{NonDet\ \ \ \ \ \ List<Det\ \ \ \ \ \ \ \ \ Integer>}
    \item \codeid{OrderNonDet List<NonDet\ \ \ \ \ \ Integer>}
    \item \codeid{Det\ \ \ \ \ \ \ \ \ List<NonDet\ \ \ \ \ \ Set<...>\relax >}
    \item \codeid{Det\ \ \ \ \ \ \ \ \ List<OrderNonDet Set<...>\relax >}
\end{itemize}

\begin{figure}
    \centering
    \begin{tabular}{|l|l|l|l|l|}
        \cline{3-5}
        \multicolumn{2}{c|}{~}  &  \multicolumn{3}{c|}{element type} \\ \cline{3-5}
        \multicolumn{2}{c|}{~}  & NonDet     & OrderNonDet & Det \\ \hline
        & NonDet      &   valid    &  invalid    & invalid  \\ \cline{2-5}
        Collection    & OrderNonDet &   invalid  &  valid      & valid  \\ \cline{2-5}
        & Det         &   invalid  &  invalid    & valid      \\ \hline
    \end{tabular}
    \caption{Valid Collection declarations.  The Collection's type qualifier
        must be a supertype or equal to the element type.}
    \label{fig-determinism-collections}
\end{figure}

A collection or map annotated as \<OrderNonDet> has the following properties:
\todo{Write these as type rules.}
\begin{enumerate}
    \item The individual elements extracted by iterating over the collection get the type \<NonDet>.
    \item Accessing properties such as \<size()> or querying if the collection \<isEmpty()> will return types
    that are annotated as \<Det> as they do not depend on the iteration order. 
\end{enumerate}

@@

Some types are invalid and must not be permitted (during polymorphic instantiation).

Recall that \theDeterminismChecker does not permit \<OrderNonDet> on a type
that isn't a collection.
As a consequence, resolution of polymorphic type qualifiers at method signatures could have undesirable effects.
Suppose a method declaration is annotated as \codeid{PolyDet int size(PolyDet List<T> list)}. Invoking this method
with an argument of type \codeid{OrderNonDet List<Det String>} will result in \<PolyDet> getting resolved to \<OrderNonDet> which is invalid
on the return type \<int>. We circumvent this scenario by introducing two variants to the \<PolyDet> qualifier, \<PolyDet$\uparrow$>
and \<PolyDet$\downarrow$>.

@@

\begin{figure}
    $\infer[\textsc{poly resolution}]{\pi_r \ \beta_r\ \<m>(\overline{\pi_i \ \beta_i})\{\ \}}{\forall \kappa \in \{\|NonDet, OrderNonDet, Det|\} : \kappa \ \beta_r\ \<m>(\overline{K \ \beta_i})\{\ \} & \pi_r, \pi_i \in \{\|PolyDet|, \|PolyDet|\uparrow, \|PolyDet|\downarrow    \}}$
    \caption{Polymorphic resolution rule}
    \label{fig-poly-resolution}
\end{figure}

\begin{table}[]
    \begin{tabular}{|l|l|l|}
        \hline
        \textbf{\<PolyDet>} & \textbf{\<PolyDet$\uparrow$>} & \textbf{\<PolyDet$\downarrow$>} \\ \hline
        \<NonDet> & \<NonDet> &  \<NonDet>\\ \hline
        \<OrderNonDet> & \<NonDet> &  \<Det>\\ \hline
        \<Det> & \<Det> &  \<Det>\\ \hline
    \end{tabular}
\caption{Polymorphic resolution}
\todo{Write this as a rule rather than a table.}
\label{tab-poly-resolutions}
\end{table}

@@

In order to provide soundness guarantees, we must ensure that a collection that is
declared as \<Det> is not side-effected in non-deterministic ways. Consider the following qualifier 
on the List add method: \codeid{public void add(PolyDet ArrayList<E> this, PolyDet int index, E element)}.
Suppose a client calls this method as follows:
\begin{verbatim}
@Det ArrayList<@Det String> list;
@NonDet int random;
@Det String str;
list.add(random, str);
\end{verbatim}
This will result in the instantiation \codeid{public void add(@NonDet ArrayList<E> this, @NonDet int index, E element)} and
the method invocation will type check. This is problematic because even though \<list> was declared to be \<Det>,
the type checker allows the addition of an element at a \<NonDet> location violating our determinism guarantees.
We prevent this behavior by introducing another variant of \<PolyDet>, \<PolyDet("use")> that does not affect
type instantiations. This is especially useful in preventing methods from non-deterministically modifying the state
of a deterministic receiver. To have this effect, the library method \<ArrayList.add> is annotated as:
\codeid{public void add(PolyDet ArrayList<E> this, PolyDet("use") int index, E element)}.
Since \<PolyDet("use")> has no effect on type instantiations, calling this method on a \<Det List>
will result in the method declaration being resolved as \codeid{public void add(@Det ArrayList<E> this, @Det int index, E element)}
irrespective of the type of \<index>, thereby preventing undesired side-effects.

@@

With the above rules, the checker framework will automatically annotate a local collection with type \<NonDet> and its
type arguments will get the type qualifier \<Det>. For example, a locally declared \codeid{List} of \codeid{Strings} will be automatically annotated
as \codeid{NonDet List<Det String>}. Recall from~\ref{collection-rules} that \theDeterminismChecker does not allow this type on
any collection. While this is still sound, it will result in \theDeterminismChecker reporting a lot of errors because every locally declared 
collection gets annotated with this invalid type. To avoid this situation, we make an exception to the defaulting rules for local collections
and annotate their type parameters also with \<NonDet>. So a local list of Strings gets the default qualifier of \codeid{NonDet List<NonDet String>}. While this defaulting eliminates the possibility of automatically annotating locals collections with an invalid
type, it could result in a high number of false positives. In the following code snippet,
\begin{verbatim}
void test(@Det List<@Det String> argList) {
    List<String> localList = argList;
}    
\end{verbatim}
\theDeterminismChecker would report an error at the assignment statement since the inferred type for \codeid{localList} is 
\codeid{NonDet List<NonDet> String} and since collections types are invariant.
We therefore recommend that programmers explicitly annotate
all local variables that are collections to reduce false positives.

\TheDeterminismChecker has additional type refinement rules for the sorting
methods such as \<List.sort>, \<Collections.sort>, and \<Arrays.sort>:
 as shown in Figure~\ref{fig-sorting}. 
\begin{figure}
%     $\infer[\textsc{list sort}]{\Gamma \vdash x : \|Det \ List|\langle\|Det| \ \beta\rangle}{\Gamma \vdash x : \|OrderNonDet \ List|\langle\|Det| \ \beta\rangle & \|x.sort()|}$
%     
%     \bigskip

    $\infer[\textsc{sort}]{\Gamma \vdash x : \|Det \ List|\langle\|Det| \ \beta\rangle}{\Gamma \vdash x : \|OrderNonDet \ List|\langle\|Det| \ \beta\rangle & \|Collections.sort(x)|}$

%     \bigskip
%     
%     $\infer[\textsc{arrays sort}]{\Gamma \vdash x : \|Det \ \beta \ Det[]| }{\Gamma \vdash x : \|Det \ \beta \ OrderNonDet[]| & \|Arrays.sort( \ x)|}$

    \caption{Sorting type refinement rules}
    \label{fig-sorting}
\end{figure}
\TheDeterminismChecker type refines a receiver of type \codeid{OrderNonDet List} to a \codeid{Det List} if all the
type argument of this \codeid{List} have the type qualifier \codeid{Det} (Rule LIST SORT in Figure~\ref{fig-sorting}).
As per these rules, if \codeid{sort()} is called on a receiver of type \codeid{OrderNonDet List<Det Set<Det String>>}, it will
be type refined to \codeid{Det List<Det Set<Det String>>}. However a receiver of type \codeid{OrderNonDet List<OrderNonDet Set<Det String>>} will not be type refined.

@@

In the JDK, \<LinkedHashSet> is implemented as a subclass of \<HashMap>.
\todo{Talk about this design choice. Josh Bloch's book: "If the code surrounding the first declaration depended on LinkedHashSet's ordering policy,
    then it would be incorrect to substitute HashSet for LinkedHashSet in the declaration, because HashSet makes no guarantee concerning iteration order."}

\subsection{Determinism checker specific rules for precision}\label{precision}
We annotated the \codeid{equals()} method of the \<Set> class in the JDK as
\begin{verbatim}
PolyDet$\uparrow$ boolean equals(PolyDet Set<E> this, PolyDet Object o)
\end{verbatim}
While this is sound, it is imprecise in the case of \codeid{OrderNonDet Sets}. We added the following rule shown in Figure~\ref{fig-precision-rules} to our checker to improve precision.
\begin{figure}
    \todo{add line break}
    $\infer[\textsc{set precision}]{\Gamma \vdash \|z : Det|}{\Gamma \vdash \|x| : \|OrderNonDet| \ \|Set|\langle \kappa_{x} \ \beta_{x}\rangle, \|y| : \|OrderNonDet| \ \|Set|\langle \kappa_{y} \ \beta_{y}\rangle & \|z = x.equals(y)| & \langle \kappa_{x} \ \beta_{x}\rangle \nsqsubseteq \langle \|OrderNonDet \ List| \rangle & \langle \kappa_{y} \ \beta_{y}\rangle \nsqsubseteq \langle \|OrderNonDet \ List| \rangle }$

@@

    \caption{Improving precision}
    \label{fig-precision-rules}
\end{figure}
According to this rule, if the \<Set> receiver has type \codeid{@OrderNonDet Set<@Det List} and the argument to the \codeid{equals()} method is
also of the same type, the return type gets the qualifier \<Det>. Without this precision rule, the return type would have resolved to \<NonDet> which
is sound but imprecise. This rule is careful enough to not annotate the return type of \<equals()> as \<Det> if a type argument to either the
receiver or the argument happens to be an \codeid{OrderNonDet List} (An order nondeterministic list does not provide any guarantee on the
equality as it is dependent on the iteration order).

@@

% LocalWords:  basetypes NonDet OrderNonDet Det Iterable formedness param1
% LocalWords:  PolyDet boolean param2 arg1 arg2 localList entrySet keySet
% LocalWords:  LinkedHashMap TreeMap HashSet LinkedHashSet LinkedHashSet's
